---
title: "convenience Heatmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cHeatmap-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

cHeatmap is a wrapper of the excellent [`ComplexHeatmap::Heatmap()`](https://jokergoo.github.io/ComplexHeatmap/reference/Heatmap.html) function with additional functions and more friendly interface for some common tasks in my work, thus called the **c**onvenience **Heatmap**. I highly recommend reading the [ComplexHeatmap book](https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html) for more advanced use.

Here are the features:

1.  Automatic or manual exclusion of outliers in the color mapping so
    that the color scale of the heatmap and annotations is not dominated by outliers.

2.  The option to set the color-value mappings for the body and annotations of heatmaps.

3.  Automatic coloring of the dendrogram.

4.  Easy highlight or display of the values of certain cells in the body and annotations of the heatmap.

5.  Discrete color-value mapping for integer matrices containing few unique values.

6.  Interface to plot across rows.

7.  Handling of edge cases:

    -   `Inf` and `-Inf` values in input matrix cause errors in `stats::dist()` for clustering, they are reset as NA.

    -   If missing values are present in the input matrix or row and column annotations, a legend for missing value is added.
    
    -   Plot a heatmap of a data.frame with columns of different data types clustered by one of the columns.


## Examples

### Exclusion of outliers in the color mapping

Outliers in a numeric matrix dominate the color scale and make the differences among many cells barely visible, such as body and `ra1` annotation below.

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = TRUE
)
```

```{r, fig.keep = 'last', fig.asp = 5/7 }
library(cHeatmap)

set.seed(100)
nMat1 <- rnorm(27)
mat1 <- matrix(c(-5,-Inf,nMat1,Inf), nrow = 6)
rownames(mat1)=paste0('r',1:nrow(mat1))
colnames(mat1)=paste0('c',1:ncol(mat1))
raDf=data.frame(ra1=c(20,Inf,1:4),ra2=c(NA,letters[1:5]))
rownames(raDf)=rownames(mat1)
caDf=data.frame(ca1=c(9.8,11.3,6.9,5.4,-80))
rownames(caDf)=colnames(mat1)

cHeatmap(
  mat1,
  NA.color = 'grey50',
  name = "value", # title of the legend for heatmap body
  rmLegendOutliers = F, # keep outliers in the legends
  cluster_rows = F, cluster_columns = F, # no clustering
  cellFun = function(x) { x }, # display all the values in the body of the heatmap
  rowAnnoDf = raDf, # add row annotation
  colmAnnoDf = caDf, # add row annotation
  annoCellFunList = list(ra1=\(x) {1:nrow(raDf)}) # display all values of ra1 annotation
)
```

Set `rmLegendOutliers = T`, default for numerical vectors, to remove outliers in the color-value mapping of the legends, so that the difference in the body and annotation can be easily visualized. See [Reference Manual](https://blueskypie.github.io/cHeatmap/reference/cHeatmap.html) for details. Note that the affected labels of the legends now prefixed with `>` or `<`, unless `mat1` is a discrete integer vector like `ra1`, i.e.  the number of unique items is less than `intAsDiscreteCutoff`.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(
  mat1,
  name = "value", 
  rmLegendOutliers = T, # remove outliers in the legends
  cluster_rows = F, cluster_columns = F, 
  cellFun = function(x) { x }, 
  rowAnnoDf = raDf, 
  colmAnnoDf = caDf,
  annoCellFunList = list(ra1=\(x) {1:nrow(raDf)},
                         ca1=\(x) {1:nrow(caDf)}) 
)
```


### Manually set the color-value mapping

Instead of auto-detecting the outliers, use the `colMap` to set the range. Values outside of the range are treated as outliers. 

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(
  mat1,
  name = "value", 
  # manually set the color mapping for the heatmap body, here only the numeric 
    # values are set since the default colors are green, white, and red.
    # one can also set only the upper or low bound, e.g. colMap = c(NA, 0, 1)
  colMap = c(-1, 0, 1), 
  rmLegendOutliers = T, # remove outliers in the legends
  cluster_rows = F, cluster_columns = F, 
  colmAnnoDf = caDf,
  # manually set the color mapping for the ra1
  colmAnnoColMap = list(ca1=c('white'=0,'gold'=12))
)


```


Set multi-color mapping to visualize local clusters or outliers 


```{r, fig.keep = 'last', fig.asp = 5/7 }
# create an Inf value
nMat1 <- rnorm(30)
uMat1 <- runif(12, -5, 20)
mat2 <- matrix(c(nMat1, uMat1), nrow = 6)
rownames(mat2)=paste0('r',1:nrow(mat1))
caDf2=data.frame(ca2=c(9.8,11.3,6.9,5.4,-80, 7, 60))
rownames(caDf2)=colnames(mat2)

bCm=c("green4" = -1, "white" = 0, "red" = 1, #cluster 1
             "yellow" = 3, "blue" = 18) #cluster 2
caCm = c('purple'=-80,'white'=0,'yellow4'=12, 'red'=60)

cHeatmap(mat2,
  name = "value",
  rmLegendOutliers = F,
  colMap = bCm,
  column_split = c(rep("nMat1", 5), rep("uMat1", 2)),
  rowAnnoDf = raDf, 
  # manually set the color mapping for the ra1
  rowAnnoColMap = list(ra1=c('white'=1,'gold'=4,'pink4'=20)),
  colmAnnoDf = caDf2,
  colmAnnoColMap = list(ca2=caCm)
  
)
```

Set distinct color to different numerical ranges by adding a `|` between the names of two colors, 
e.g. the `bCm` and `caCm` below. And adjust the legend to

-   lengthen the portion for cluster 1
-   shorten the portion between 1 and 3 if majority of cluster 2 are in (3, 18)
-   increase the height of the whole legend to 4cm

```{r, fig.keep = 'last', fig.asp = 5/7 }
bAt=c(-1,0,1,3,6,9,12,15,18)
bCm=c(green = -1,
           'green4|yellow' = 0,
           'yellow4|red'=1,
           'red4|deepskyblue' = 3,
           blue4 = 18)
caCm= c('purple'=-80,'purple2|white'=3,'yellow4|red'=12, 'red2'=60)
caAt=c(-80,3,6,9,12,60)
cHeatmap(mat2,
         name = "value",
         rmLegendOutliers = F,
         #set color mapping with distinct regions
         colMap = bCm, 
         column_split = c(rep("nMat1", 5), rep("uMat1", 2)),
         legendTicks = bAt,
         legendBreakDist = rep.int(1,length(bAt)-1),
         legendHeight = 4,
         rowAnnoDf = raDf,
         rowAnnoColMap = list(ra1=c('white'=1,'gold'=4,'pink4'=20)),
         colmAnnoDf = caDf2,
         colmAnnoColMap = list(ca2=caCm),
         colmAnnoPara = list(annotation_legend_param=list(
           ca2=list(at=caAt,break_dist=rep.int(1,length(caAt)-1))
           ))
)

```



### Display cell values using `cellFun` and `annoCellFun`

Show only values in `(0.5, 1)` in the heatmap body and the 5th value in `ra1` and value `b` in `ra2`.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(
  mat1,
  name = "value", 
  cellFun = \(x) { if (is.finite(x) && x > 0.5 && x < 1) x  }, 
  rowAnnoDf = raDf, 
  annoCellFunList = list(ra1=\(x) {5},
                         ra2=\(x) {which(x=='b')}) 
)
```

Display only the outliers in the heatmap body and ra1.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(
  mat1,
  name = "value", 
  cellFun = 'o', 
  rowAnnoDf = raDf, 
  annoCellFunList = list(ra1=\(x) 'o') 
)
```

Mark outliers as `X`; 

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(
  mat1,
  name = "value", 
  cellFun = c("o", "X"), 
  rowAnnoDf = raDf, 
  annoCellFunList = list(ra1=\(x) list('o','X')) 
)
```

Color outliers by black edge.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(mat1, 
  cellFun = c(
    "o", list("rect", col = "black", lwd = 2)
  )
)
```

Display `H` if cell values \> 2 and `L` if \< 0.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(mat1,
  cellFun = function(x) {
    if(is.finite(x)){
      if(x > 2) "H" else if(x < 0) "L"
    }
  }
)
```

Add black edge if cell values \> 2 and display `L` if \< 0.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(mat1,
  cellFun = function(x) {
    if(is.finite(x)){
      if(x > 2) list("rect", col = "black", lwd = 2) else if(x < 0) "L"
    }
  }
)
```

### Color dendrograms and split the clusters

`nRowCluster` and `nColmCluster` specify the number of colors in the corresponding dendrograms.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(mat1,
  name = "value",
  rmLegendOutliers = F,
  nRowCluster = 2, nColmCluster = 3,
  row_split = 2, column_split = 3
)
```

### Integer matrices

If the number of unique values is greater than `intAsDiscreteCutoff` whose default value is 6 in `cHeatmap` parameter settings, the color mapping is continuous; otherwise, the mapping is discrete.

```{r, fig.keep = c(1,3), fig.asp = 5/7 }
mati <- matrix(sample(1:100, 42, TRUE), nrow = 6)
cHeatmap(mati, name = "value")

mati <- matrix(sample(1:5, 42, TRUE), nrow = 6)
cHeatmap(mati, name = "value")
```

Set colors manually.

```{r, fig.keep = 'last', fig.asp = 5/7 }
cHeatmap(mati,
  name = "value",
  colMap = c("red" = 1, "pink" = 2, "yellow" = 3, "green" = 4, "green4" = 5)
)
```



### Concatenate two heatmaps

Set `drawHeatmap = F` to concatenate multiple heatmaps.

```{r, fig.keep = 'last', fig.height=7,fig.width=10.5}
nMat1 <- matrix(rnorm(30), nrow = 6)
uMat1 <- matrix(runif(12, -5, 20), nrow = 6)

hm1 <- cHeatmap(nMat1, drawHeatmap = F, name = "nMat1")
hm2 <- cHeatmap(uMat1, drawHeatmap = F, name = "uMat1")
hm1 + hm2
```

### Plot inside each row using `rowDraw`

Here `painScore` of `patient 2` is plotted across visits. Its values are transformed to the range of (0,1), representing relative values across visits.

```{r, fig.keep = 'last', fig.height=7,fig.width=9.3,dpi=100 }
mat1 <- matrix(c(nMat1, uMat1), nrow = 6)
rownames(mat1) <- paste("pt", 1:6)
colnames(mat1) <- paste("visit", 1:7)

painScore <- rnorm(7)

cHeatmap(mat1,
  name = "value", cluster_columns = F,
  rowDraw = list(
    list("grid.lines", col = "black", lwd = 2),
    matrix(painScore, nrow = 1), # data to be plotted
    2 # plot at the 2nd row of mat1
  )
)
```

Plot both lines and points in multiple rows.

```{r, fig.keep = 'last', fig.height=7,fig.width=9.3,dpi=100}
# longitudinal scores of four patients
painScore <- matrix(rnorm(28), nrow = 4)
cHeatmap(mat1,
  name = "value", cluster_columns = F,
  rowDraw = list(
    list(
      list("grid.points", size = 0.5, pch = 15, col = "blue"),
      list("grid.lines", col = "black", lwd = 2)
    ),
    painScore, 
    c(2, 3, 4, 5) # row indices of the four patients in mat1
  )
)
```

### Plot a heatmap of each column of a data.frame of different data types

```{r, fig.keep = 'last', fig.asp = 5/7 }
df2=cbind(m1=mat1[,1],raDf)
cHeatmap(
  df2[1],
  name = "m1", 
  rowAnnoDf = df2[2:3]
)
```
